---
title: "The TEACH Intervention for Youth with Childhood-Onset Lupus"
subtitle: "Research Case Study for OST 597 (Biomedical Research Structure and Methods)"
author: 
  - name: Steven J. Pierce
    orcid: 0000-0002-0679-3019
    email: pierces1@msu.edu
    affiliations: 
      - name: Michigan State University, Center for Statistical Training and Consulting
  - name: Natoshia R. Cunningham
    orcid: 0000-0002-1788-7264
    affiliations: 
      - name: Michigan State University, Department of Family Medicine
  - name: Andrea Knight
    orcid: 0000-0003-4278-9049
    affiliations: 
      - name: University of Toronto, Temerty Faculty of Medicine
date: March 30, 2026
title-slide-attributes: 
  data-notes: 1 hr (40-45 min talk + 5 min Q & A + 10 min break) My name is 
              Steven Pierce. I'll be presenting a research case study on the 
              TEACH intervention for youth with childhood-onset lupus. I was 
              the statistician on the project, so my talk will emphasize some of 
              the research design and methodology issues involved, plus the 
              results we obtained.  
csl: apa-numeric-superscript-brackets.csl      # File controls citation & reference list format
bibliography: references.bib                   # File holds BibTeX data for references
format: 
  revealjs:                                    # Settings for HTML revealjs slide output
    embed-resources: true                      # Makes HTML output file self-contained
    theme: [default, brand, CSTAT_theme.scss]  # Theme controls output formatting
    controls: true                             # Turn on visible slide controls
    slide-number: c/t                          # Turn on slide numbering
    show-notes: false                          # Hide notes (except in speaker view)
    header-logo: graphics/Combomark-Horiz_Green-RGB.png
    footer: <a href="https://github.com/sjpierce/OST597.TEACH">https://github.com/sjpierce/OST597.TEACH</a>
    link-external-icon: true                   # Mark links that lead to external sites
    link-external-newwindow: true              # Open external web links in a new window
filters:                                       # Filters are Quarto extensions
  - reveal-header                              # Allows footer text & header logo
---

```{r}
#| label: load-packages
#| include: false

library(here)             # for here()
library(tidyverse)        # loads dplyr, ggplot2, lubridate, stringr all at once.
                          # for %>%, filter(), mutate(), etc.
library(knitr)            # for kable(), include_graphics()
library(kableExtra) # Functions for formatting nice tables. 
library(emmeans)          # for emmeans(), contrast()
library(mice)             # for mice(), flux(), etc. 
library(miceadds)         # for mi.anova()
```

``` {r}
#| label: load-OST597-data
#| include: false

# Data file created by 
# P:\Consulting\Cases_1400-1599\C1563\TEACHPilot\scripts\Outcome_Analyses.qmd
# and copied from 
# P:\Consulting\Cases_1400-1599\C1563\TEACHPilot\data\OST597_TEACH.RData

load(file=here("data/OST597_TEACH.RData"))
```

## Preface

* Presentation is based on a forthcoming paper. [@Cunningham-RN8694]
* See the paper for most supporting references. 
* Study was approved by MSU IRB (STUDY00003882).

:::{.callout-tip title="Funding"}
All phases of this study were supported by a Childhood Arthritis and 
Rheumatology Research Alliance-Arthritis Foundation Transdisciplinary Research 
Grant (PI: Natoshia R. Cunningham).

:::

::: {.notes}
My case study is based on a forthcoming paper in *Arthritis Care & Research*. 
Supporting references for some of my points here can be found in the paper, 
which is listed in the references at the end of the slides. I've just been 
selective about citing sources in the slides to keep them cleaner. The study was 
approved by MSU's institutional review board. 
:::

## Contributors {.smaller}

:::: {.columns}
::: {.column width="34%"}
* **Natoshia R. Cunningham**
* Thea Senger-Carpenter
* Jocelyn Zuckerman
* Michelle Adler
* Mallet R. Reid 
* Ashley Danguecan
* Luana Flores Pereira
* Sarah I. Mossad
* Samantha L. Ely
* Khalid Abulaban
:::

::: {.column width="33%"}
* Elizabeth A. Kessler
* Natalie Rosenwasser
* Tamar B. Rubinstein
* Ekemini A. Ogbu
* Emily A. Smitherman
* Alaina Miller
* Taylor Abounader
* Elizabeth Ross
* Livie Timmerman
* Dhriti Sharma
:::

::: {.column width="33%"}
* Jennifer N. Stinson
* Stacy Allen
* Kabita Nanda
* Tala El Tal
* Deborah M. Levy
* Linda T. Hiraki
* Hermine I. Brunner
* Mathew Reeves
* Steven J. Pierce
* **Andrea Knight**
:::

:::: 

**Bold = Principal investigators.**

::: {.notes}
Our study team was very large. This is a list of all the co-authors on our paper
with the principal investigators' names in bold text. That said, let's get
started with the substantive content of the case study.
:::

# Introduction

::: {.notes}
:::

## Childhood-Onset Systemic Lupus Erythematosus

* cSLE is a chronic, multisystemic, autoimmune disease diagnosed prior to age 18 
  years.
* Higher risk for depressive, fatigue, and pain symptoms, which may lower 
  health-related quality of life (HR-QOL).
* More often affects female and Asian, Native American, and Black youth.
* Black youth may get less help managing comorbid symptoms than White peers.

::: {.notes}
Childhood-onset lupus is a chronic, multisystemic, autoimmune disease diagnosed 
prior to age 18 years. Patients are at higher risk for depressive, fatigue, and 
pain symptoms, which may lower health-related quality of life. We know that 
lupus has a disproportionate impact on some parts of the population: it is more 
likely to affect female and Asian, Native American, and Black youth. It's the 
fifth leading cause of death for Black and Hispanic females age 15-24. Previous
studies also suggest that Black youth may be less likely to receive effective 
help managing comorbid symptoms such as anxiety and depression than White peers.
This all means that we need better interventions to help lupus patients. 
:::

## Cognitive Behavioral Therapy (CBT)

* Effective for managing depression, fatigue, and pain in youth with other 
  chronic illnesses and in *adults* with lupus. 
* Prior studies on CBT for *youth* with lupus: 
    * One failed to find clinically significant effects on pain, affect, & 
      quality of life. 
    * Treatment and Education Approach for Childhood-Onset Lupus (TEACH) program 
      had promising results but needed refinements. 

::: {.notes}
Psychological interventions offer one way to improve the mental health of lupus
patients. Research has shown that cognitive behavioral therapy is effective for
managing depression, fatigue, and pain in youth with other chronic illnesses and
in adults with lupus. However, there were only two prior studies on using CBT
for youth with lupus. One of them failed to find any clinically significant
effects on pain, affect, and quality of life. The other was our team's
single-arm pilot test of the Treatment and Education Approach for
Childhood-Onset Lupus, the TEACH program. My colleagues' first TEACH study 
yielded promising results but needed some refinements.
:::

## TEACH: Initial Pilot Study

* A tailored, CBT program delivered via 6 weekly sessions.
* Targets depressive, fatigue, & pain symptoms as primary outcomes.
* High retention rate (82.4%).
* Reduced depressive and fatigue symptoms.
* Marginal reductions in pain and anxiety. 

::: {.notes}
The first iteration of the TEACH intervention consisted of a tailored, CBT 
program delivered during a series of 6 weekly in-person visits. The primary 
outcomes it sought to improve were patients' depressive, fatigue, and pain
symptoms. The first study showed a retention rate of just over 82% and potential 
improvements in depressive and fatigue symptoms, with weaker evidence for 
reductions in pain and anxiety. 

Intervention development is often iterative. Early results like these are 
promising but not definitive and usually accompanied by feedback about ways to 
improve the program. 
:::

## TEACH: Initial Lessons Learned

* Patients wanted a remote delivery option. 
* Need to engage patients in refining the intervention.
* Need to better address pain and anxiety. 
* Need to get stronger evidence of efficacy.

::: {.notes}
Some of the initial lessons learned were that patients reported wanting the
option to get the TEACH intervention delivered remotely instead of only through
in-person visits. That first study also revealed a need to refine the
intervention and it was clear that engaging lupus patients in planning those
refinements could be crucial to improving feasibility, acceptability, and
effectiveness of the intervention. One priority was better addressing pain and
anxiety. Finally, my colleagues also recognized that they needed a stronger
research design because the initial study lacked a control group. Moving to a
randomized trial could yield stronger evidence for TEACH's efficacy.
:::

## Current Study

* Update program content per previous feedback.
* Examine feasibility & acceptability of remotely-delivered TEACH for 
  adolescents/young adults w/ cSLE. 
* Test efficacy of TEACH vs. TAU:
    * Primary: depressive, fatigue, & pain symptoms. 
    * Secondary: disease activity, medication adherence, & anxiety symptoms.

::: {.notes}
That led to development of the study I'm describing today. The PIs decided to
update the program content and delivery method to address the patient feedback
from the prior study. So, our new study needed to again examine the feasibility
and acceptability of the updated program, which was delivered remotely via
video-conference calls with participants instead of in-person. The other major
objective was to move to a randomized controlled trial design that included an
intervention arm and a treatment-as-usual control arm. That design change aimed
to improve internal validity when we tested the intervention's efficacy by
comparing primary and secondary outcomes among patients who received TEACH
with outcomes among those who did not. Both groups got medical
treatment-as-usual for lupus throughout the study, so study arms were a
TEACH+TAU intervention arm and TAU-alone control arm.

Measures of depressive, fatigue, and pain symptoms were the primary outcomes, 
while measures of disease activity, medication adherence, and anxiety symptoms 
were secondary outcomes. 
:::

# Methods

::: {.notes}
:::

## Research Design

```{r}
#| results: asis

StudyDesign <- data.frame(Arm = c("TEACH", "TAU"),
                          Pretest = rep("$O_1$", times = 2),
                          Allocation = rep("$R$", times = 2),
                          Intervention = c("$X$", ""),
                          Posttest = rep("$O_2$", times = 2)) 

FN <- paste("$O_i$, observed data at time $i$;", 
            "$R$, randomized;",
            "$X$, delivery of TEACH intervention (6 weeks).")
kable(StudyDesign, format = "html", escape = FALSE, align = "lcccc") %>% 
  kable_styling(full_width = FALSE) %>% 
  add_header_above(c(" ", "Longitudinal Event" = 4), escape = FALSE) %>% 
  footnote(general = FN, general_title = "Note: ", footnote_as_chunk = TRUE,
           threeparttable = FALSE, escape = FALSE) %>% 
  cat()
```

* Preregistered at ClinicalTrials.gov (NCT04335643)

::: {.notes}
This international, multi-site study used a parallel arms randomized controlled 
trial design, with pretest data collected before random assignment. The TEACH 
arm got the intervention between pretest and posttest, but the TAU arm did not.

:::

## Sites: Pediatric Rheumatology Clinics

:::: {.columns}
::: {.column width="50%"}
**United States**

* Helen DeVos Children's Hospital
* Cincinnati Children's Hospital Medical Center
* Seattle Children's Hospital
* Children's Hospital at Montefiore
* University of Alabama

:::

::: {.column width="50%"}
**Canada**

* The Hospital for Sick Children, Toronto

:::
::::

::: {.notes}
Our sites were all pediatric rheumatology clinics. We had five sites in the US 
and one in Canada. 
:::

## Participants {.smaller}

:::: {.columns}
::: {.column width="50%"}
**Inclusion Criteria**

* Age 12-22 years, 
* cSLE diagnosis by age 18 yrs,
* Elevated depressive, fatigue, and/or pain symptoms, and
* English language proficiency. 

:::

::: {.column width="50%"}
**Exclusion Criteria**

* Other chronic medical conditions, 
* Developmental delays or severe cognitive impairments,
* Untreated major psychiatric illness, or
* In treatment for depressive, fatigue, or pain symptoms. 

:::
::::

::: {.notes}
Participants had to be 12-22 years old, with a cSLE diagnosis from a pediatric 
rheumatologist by age 18 yrs, have elevated depressive, fatigue, and/or pain 
symptoms, and proficient in English. We excluded any who had other chronic 
medical conditions (such as juvenile idiopathic arthritis), any documented 
developmental delays or severe cognitive impairments, untreated major 
psychiatric illness (such as bipolar disorder, active suicidal ideation), or
who was already getting treated for depressive, fatigue, or pain symptoms. 

Participants were compensated with gift cards, with $50 each at baseline and and 
post-assessment. 
:::

## TEACH Intervention

* Advisory input from cSLE patients & caregivers.
* A tailored, CBT program, 6 weekly video calls.
* Delivered by 1 of 3 interventionists.
* Supporting online modules, caregiver involved for minors
* Still targets depressive, fatigue, & pain symptoms.
* Psychoeducation, behavioral, cognitive, & mindfulness strategies. 

::: {.notes}
The TEACH intervention was adapted with input from an advisory group of lupus 
patients and caregivers. It remained a 6-session CBT program, but was delivered 
via video calls with trained interventionists (two licensed clinical social 
workers or a pediatric psychologist). Patients used supporting online modules, 
and caregivers were involved for minors. TEACH continued to target depressive, 
fatigue, and pain symptoms as primary outcomes. The intervention content 
included psychoeducation, behavioral stratigies such as relaxation training and 
activity pacing, cognitive strategies such as challenging automatic thoughts and 
problem solving, and mindfulness strategies. 
:::

## Fidelity Assessment

* All sessions video recorded.
* Independent evaluators scored fidelity 0-100% via standardized checklist. 
* Average fidelity was 98%

::: {.notes}
The team recorded video of all intervention sessions, then had three 
independent, trained evaluators (a physician, licensed social worker, or
doctoral student of psychology) use standardized checklists to assess fidelity 
to the TEACH intervention manual. Fidelity was extraordinarily high: the average 
across all sessions of 98%, so the intervention was delivered as designed. 
:::

## Stratified Random Block Randomization

```{r}
#| results: asis

Strata <- data.frame(Stratum = 1:4,
                     Country = rep(c("United States", "Canada"), each = 2),
                     Age_Group = rep(c("Adolescent (12-17 yrs)", 
                                       "Young Adult (18-22 yrs)"), times = 2)) 

FN <- paste("Within stratum, 1:1 ratio for arms (random blocks, n = 4 or 8).",
            "Investigators were blind to randomization.",
            "TAU arm offered TEACH after posttest.")
kable(Strata, format = "html", escape = FALSE, align = "ccc",
      col.names = c("Stratum", "Country", "Age Group")) %>% 
  kable_styling(full_width = FALSE) %>% 
  footnote(general = FN, general_title = "Note: ", footnote_as_chunk = TRUE,
           threeparttable = FALSE, escape = FALSE) %>% 
  cat()
```

::: {.notes}
To balance sample size across arms, the study used a stratified random block
randomization process. The two stratification variables were country (US vs
Canada) and age group (adolescent vs young adult). Within each stratum, blocks
of either 4 or 8 participants were allocated to the TEACH vs TAU arms in 1:1
ratio. The investigators (except the interventionists & statistician) were blind
to randomization results during data collection. After the posttest, the TAU
participants were offered the TEACH intervention.
:::

## Primary Outcome Measures 

* Recruitment rate = (# consented)/(# approached)
    * Benchmark: > 65 %
* Retention rate = (# completed posttest)/(# randomized)
    * Benchmark: > 80%
* Qualitative interviews about TEACH acceptability

::: {.notes}
The primary quantitative outcomes for this study were the recruitment rate, 
defined as the percent of approached patients who consented to joining the 
study, and the retention rate, defined as the percent of randomized patients who 
completed the posttest data collection. In addition, qualitative feedback about 
the acceptability of the intervention was collected by interviewing 
participants.
:::

## Secondary Outcome Measures

* Depressive symptoms over past 2 weeks
    * Children’s Depression Inventory 2nd Edition (CDI-II) for adolescents, 28 items
    * Beck Depression Inventory-II (BDI-II) for young adults, 21 items
    * Both converted to $T$-scores (M = 50, SD = 10) 
    * Eligible if $60 \le T < 90$.
    * 50% reduction is meaningful.
* Fatigue over past week
    * PROMIS Pediatric Short Form for adolescents, 10 items
    * PROMIS Adult Fatigue Short Form for young adults, 8 items
    * Both converted to $T$-scores (M = 50, SD = 10) 
    * Eligible if $T \ge 60$ (elevated).
    * Reduction of $\ge 10$ is meaningful. 
* Pain intensity over past 2 weeks
    * Pain Visual Analog Scale (VAS), ranges 0-10
    * Eligible if rating $\ge 3$ (moderate)
    * Reduction of $\ge 30$% are meaningful.
  
::: {.notes}
:::

## Exploratory Outcome Measures

* Anxiety for past 3 months
    * Screen for Child Anxiety Related Disorders (SCARED) for adolescents, 41 items
    * Screen for Adult Anxiety Related Disorders (SCAARED) for young adults, 44 items
    * Converted to $z$-scores.
    * Reduction of >50% is meaningful. 
* Disease Activity
    * Systemic Lupus Erythematosus Disease Activity Index (SLEDAI-2K).
    * Scores range from 0-105, with scores $\ge 6$ indicating active disease. 
* Medication adherence over past month
    * Adapted from Medication Adherence Self-Report Inventory
    * Rated 0-100%, scores $\ge 80$% indicate adequate adherence.
* Pediatric health-related quality of life (HR-QOL)
    * PedsQL Quality of Life Inventory (child & caregiver)
    * PedsQL Rheumatology Module (child & caregiver)
* Adverse events during the intervention
    * Negative Effects Questionnaire, 20 items
    * Binary indicator of occurrent (0 = no, 1 = yes)
    * Ordinal indicator of how negatively they were affected (0-5)

::: {.notes}
:::

## Covariates

* Demographics: Age, sex, race, ethnicity, disease duration (months), education.
* Disease manifestations: Systemic Lupus International Collaborating Clinics 
  (SLICC) classification.
* Adverse childhood experiences (ACEs).
* Medication usage (name, indication). 

## Quantitative Analysis

* Estimated recruitment and retention rates
* Intent-to-treat (ITT) estimate of intervention effect
* ANCOVA predictors or posttest outcomes: 
    * Arm (focal predictor)
    * Pretest outcome
    * Country
    * Age group
    * Covariates: Race, ethnicity, ACE exposures
    * Arm x pretest
* Estimated marginal means for arm effect

::: {.notes}
:::

## Imputing Missing Data via MICE {.smaller}

* Allowed retaining all cases as required for ITT analysis 
* 55% of cases had at least one missing value.
    * HR-QOL measures missing by design for young-adults. 
* Imputation model used:
    * All focal predictors & covariates used in the (ANCOVA) model.
    * Binary indicator of post-assessment attendance.
    * Longitudinal relationships between pretest & posttest for each outcome.
    * Cross-sectional relationships between outcomes measured at the same time. 
* Parameter estimates pooled across analyses of 55 imputed datasets. 

::: {.notes}
To ensure that we retained all randomized participants in the analysis while 
also handling missing data, we used multiple imputation via chained equations 
:::

# Results

## Sample Descriptives

* Country: 61% US, 39% CA
* Age: M = 16.5, SD = 2.1, 64% adolescent, 36% young adult
* Sex: 97% female
* 


::: {.notes}
:::

## CONSORT Chart

```{dot}
//| fig-width: 10
//| fig-height: 2.5
digraph RRProduct {
rankdir="LR";

node [shape = "box", style= "filled", fontname="Lato", fillcolor = "#C3FFEC"]
Identified [label = "Identified\nn = 293"]
Approached [label = "Approached\nn = 200"]
Consented [label = "Consented\nn = 97"]
Screened [label = "Screened\nn = 85"]
Pretest [label = "Pretest &\nRandomized\nn = 64"]
TEACH [label = "TEACH\nn = 32"]
TEACHPosttest [label = "Posttest\nn = 29"]
TAU [label = "TAU\nn = 32"]
TAUPosttest [label = "Posttest\nn = 30"]

Identified -> Approached -> Consented -> Screened -> Pretest 
Pretest -> TEACH -> TEACHPosttest
Pretest -> TAU -> TAUPosttest
}
```

:::: {.fragment}
:::{.callout-tip title = "Primary Outcomes"}

* Recruitment rate = 97/200 = 48.5% (benchmark: 65%)
* Retention rate = 59/64 = 92.2% (benchmark: 80%)
    * Unrelated: arm, country, age group, race, ethnicity, most pretest 
      outcomes. 
    * Related: pretest child rheumatology QOL, disease duration.

:::
::::

::: {.notes}
This CONSORT chart summarizes participant flow through the study. It started by 
identifying potential participants, then approaching them to ask for consent. 
Then, those who consented were screened and the eligible participants then took
the pretest and were randomly assigned to one of the two arms (TEACH vs TAU). 
Then, 8 weeks later, the posttest assessment occurred. The chart in our paper 
shows breakdowns of why people dropped out between steps. Here, I want 
to emphasize that we can use the data shown here to estimate two of the primary 
outcomes, the recruitment rate and the retention rate. 97 of 200 patients 
approached consented to participate, yielding a 48.5% recruitment rate. That was 
lower than the 65% benchmark we hoped to achieve. However, 59 of 64 randomized 
participants returned for the posttest assessment, giving us a 92.2% retention 
rate that exceeded the 80% benchmark we had set. In our attrition analyses, 
arm, country, age group, race, ethnicity, most pretest outcomes were unrelated 
to whether someone was retained at posttest, but pretest child rheumatology 
quality of life and disease duration were related to retention. 
:::

## Primary Outcomes: Acceptability Themes

* Skills were logically connected & went "hand in hand".
* Frequent use of relaxation skills, pleasant activity planning, & activity 
  pacing was helpful.
* There was immediate effect of using the skills taught.
* Positive relationship w/ interventionist motivated practicing skills.
* Symptom severity & busy schedules were barriers to using skills.
* Virtual format was appreciated and it facilitated attendance. 

::: {.notes}
:::


## Secondary Outcomes: Arm Main Effect

```{r}
#| fig-height: 1.5
#| fig-width: 5
#| layout-ncol: 2

plot(emm_mood, ylab = NULL, xlab = "Depressive symptoms, Posttest Means") +
  scale_x_continuous(limits = c(40, 70), breaks = seq(40, 70, 10))

plot(emmc_CI_mood, ylab = NULL, xlab = "Depressive symptoms, Posttest Difference") + 
  scale_x_continuous(limits = c(-15, 15), 
                     breaks = seq(from = -15, to = 15, by = 5)) +
  geom_vline(xintercept = 0, col = "red") 

plot(emm_fatigue, ylab = NULL, xlab = "Fatigue, Posttest Means") +
  scale_x_continuous(limits = c(40, 70), breaks = seq(40, 70, 10))

plot(emmc_CI_fatigue, ylab = NULL, xlab = "Fatigue, Posttest Difference") + 
  scale_x_continuous(limits = c(-15, 15), 
                     breaks = seq(from = -15, to = 15, by = 5)) +
  geom_vline(xintercept = 0, col = "red") 

plot(emm_pain, ylab = NULL, xlab = "Pain, Posttest Means") +
  scale_x_continuous(limits = c(1, 5), breaks = seq(1, 5, 0.5))

plot(emmc_CI_pain, ylab = NULL, xlab = "Pain, Posttest Difference") + 
  scale_x_continuous(limits = c(-2, 2), 
                     breaks = seq(from = -2, to = 2, by = 1)) +
  geom_vline(xintercept = 0, col = "red") 
```

::: {.notes}
Here are the results for testing the main effect of arm, which tells us whether 
on average the TEACH arm did better than the TAU arm at posttest. On the left, 
you can see the posttest means and 95% confidence intervals for each arm on each
secondary outcome. On the right, I've graphed the raw effect size measured by
the difference between those means. When the red reference line intersects the
confidence interval for the difference, that means there may be no difference. 
Here we see that on average, depressive symptoms improved by almost 8 points
and fatigue by almost 4 points. Pain, however may not have improved. 
:::

## Fatigue: Arm x Pretest Interaction

```{r}
#| fig-width: 10
#| fig-height: 5.5

emmip(fit_fatigue_m3, formula = arm ~ c_fatigue_T1, cov.reduce = range, 
      style = "numeric", CIs = TRUE, CIarg = list(lwd = 1, alpha = 0.25), 
      linearg = list(lwd = 1.2), 
      xlab = "Centered Pretest Fatigue",
      ylab = "Posttest Fatigue") + 
  geom_ribbon(aes(ymin=LCL, ymax=UCL, fill=arm), alpha=.25) + 
  theme(text = element_text(family = "sans", size = 14),
        axis.text = element_text(family = "sans", size=14),
        legend.position = "inside", 
        legend.position.inside = c(0.85, 0.20),
        legend.title = element_text(family = "sans", size=12),
        legend.text = element_text(family = "sans", size=12)) +
  coord_cartesian(xlim = c(-31, 21), ylim = c(20, 71))
```

::: {.notes}
Testing the arm by pretest interaction for fatigue showed that the size of the 
intervention effect depended on pretest levels of fatigue. TEACH has the 
strongest beneficial effect on posttest fatigue when pretest fatigue is very 
mild (> 10 points below average). The benefit disappears by about 7.5 points 
above average pretest fatigue. At unusually high levels of pretest fatigue 
(> 10-12 points above average), TEACH appears to worsen posttest fatigue. 
:::

## Exploratory Outcomes: Arm Main Effect

```{r}
#| fig-height: 1.5
#| fig-width: 5
#| layout-ncol: 2

plot(emm_anxiety, ylab = NULL, xlab = "Anxiety, Posttest Means") +
  scale_x_continuous(limits = c(-0.75, 0.75), breaks = seq(-0.75, 0.75, 0.25))

plot(emmc_CI_anxiety, ylab = NULL, xlab = "Anxiety, Posttest Difference") + 
  scale_x_continuous(limits = c(-0.75, 0.75), 
                     breaks = seq(from = -0.75, to = 0.75, by = 0.25)) +
  geom_vline(xintercept = 0, col = "red") 

plot(emm_sledai, ylab = NULL, xlab = "Disease Activity, Posttest Means") +
  scale_x_continuous(limits = c(1, 10), breaks = seq(1, 10, 1))

plot(emmc_CI_sledai, ylab = NULL, xlab = "Disease Activity, Posttest Difference") + 
  scale_x_continuous(limits = c(-4, 4), 
                     breaks = seq(from = -4, to = 4, by = 1)) +
  geom_vline(xintercept = 0, col = "red") 

plot(emm_masri, ylab = NULL, xlab = "Medication Adherence, Posttest Means") +
  scale_x_continuous(limits = c(80, 110), breaks = seq(80, 110, 10))

plot(emmc_CI_masri, ylab = NULL, xlab = "Medication Adherence, Posttest Difference") + 
  scale_x_continuous(limits = c(-10, 10), 
                     breaks = seq(from = -10, to = 10, by = 5)) +
  geom_vline(xintercept = 0, col = "red") 
```

::: {.notes}
:::

## Medication Adherence: Arm x Pretest Interaction

```{r}
#| fig-width: 10
#| fig-height: 5.5

emmip(fit_masri_m3, formula = arm ~ c_masri_score_T1, cov.reduce = range, 
      style = "numeric", CIs = TRUE, CIarg = list(lwd = 1, alpha = 0.25), 
      linearg = list(lwd = 1.2), 
      xlab = "Centered Pretest Medication Adherence", xlim = c(-81, 41),
      ylab = "Posttest Medication Adherence") + 
  geom_ribbon(aes(ymin=LCL, ymax=UCL, fill=arm), alpha=.25) + 
  theme(text = element_text(family = "sans", size = 14),
        axis.text = element_text(family = "sans", size=14),
        legend.position = "inside", 
        legend.position.inside = c(0.85, 0.25),
        legend.title = element_text(family = "sans", size=12),
        legend.text = element_text(family = "sans", size=12)) +
  coord_cartesian(xlim = c(-60, 20), ylim = c(40, 120))
```

::: {.notes}
While there was no main effect on medication adherence, we did find an arm by
pretest adherence interaction. That means the size of the intervention effect
was not constant. Here, we can see that TEACH has the strongest beneficial
effect on posttest medication adherence when pretest adherence is very low (> 20
points below average). The benefit disappears at average pretest adherence. At
higher levels of pretest adherence (> 10 points above average), TEACH may worsen
posttest adherence.
:::

## References {.scrollable}

::: {#refs}
:::

::: {.notes}
Finally, here's the list of references mentioned in the slides. 
:::

# Questions & Discussion Time
Thank you for attending!

::: {.notes}
*Time should be ~40-45 min, leaving 5 min for questions and discussion.*
:::
