---
title: "The TEACH Intervention for Youth with Childhood-Onset Lupus"
subtitle: "Research Case Study for Michigan State University course OST 597 (Biomedical Research Structure and Methods)"
author: 
  - name: Steven J. Pierce
    orcid: 0000-0002-0679-3019
    email: pierces1@msu.edu
    affiliations: 
      - name: Michigan State University, Center for Statistical Training and Consulting
  - name: Natoshia R. Cunningham
    orcid: 0000-0002-1788-7264
    affiliations: 
      - name: Michigan State University, Department of Family Medicine
  - name: Andrea Knight
    orcid: 0000-0003-4278-9049
    affiliations: 
      - name: University of Toronto, Temerty Faculty of Medicine
date: March 30, 2026
title-slide-attributes: 
  data-notes: 1 hr (40-45 min talk + 5 min Q & A + 10 min break) Good afternoon. 
              My name is Steven Pierce. I'm a statistician here at MSU's Center 
              for Statistical Training and Consulting. I collaborate with a 
              variety researchers on campus, contributing to research designs, 
              planning, conducting, and interpreting statistical analyses, and 
              coauthoring papers to help them do more rigorous research. I'll be 
              presenting a research case study on the TEACH intervention for 
              youth with childhood-onset lupus. Given my role on the project, my 
              talk will emphasize some features of the research design we used 
              and methodology decisions we had to make, plus the results we 
              obtained. 
csl: apa-numeric-superscript-brackets.csl      # File controls citation & reference list format
bibliography: references.bib                   # File holds BibTeX data for references
format: 
  revealjs:                                    # Settings for HTML revealjs slide output
    embed-resources: true                      # Makes HTML output file self-contained
    theme: [default, brand, CSTAT_theme.scss]  # Theme controls output formatting
    controls: true                             # Turn on visible slide controls
    slide-number: c/t                          # Turn on slide numbering
    show-notes: false                          # Hide notes (except in speaker view)
    header-logo: graphics/Combomark-Horiz_Green-RGB.png
    footer: <a href="https://github.com/sjpierce/OST597.TEACH">https://github.com/sjpierce/OST597.TEACH</a>
    link-external-icon: true                   # Mark links that lead to external sites
    link-external-newwindow: true              # Open external web links in a new window
filters:                                       # Filters are Quarto extensions
  - reveal-header                              # Allows footer text & header logo
---

```{r}
#| label: load-packages
#| include: false

library(here)             # for here()
library(tidyverse)        # loads dplyr, ggplot2, lubridate, stringr all at once.
                          # for %>%, filter(), mutate(), etc.
library(knitr)            # for kable(), include_graphics()
library(kableExtra) # Functions for formatting nice tables. 
library(emmeans)          # for emmeans(), contrast()
library(mice)             # for mice(), flux(), etc. 
library(miceadds)         # for mi.anova()
```

``` {r}
#| label: load-OST597-data
#| include: false

# Data file created by 
# P:\Consulting\Cases_1400-1599\C1563\TEACHPilot\scripts\Outcome_Analyses.qmd
# and copied from 
# P:\Consulting\Cases_1400-1599\C1563\TEACHPilot\data\OST597_TEACH.RData

load(file=here("data/OST597_TEACH.RData"))
```

## Preface

* Presentation is based on a forthcoming paper. [@Cunningham-RN8694]
* Slides are available from the link at bottom of screen. 
* Study was approved by MSU IRB (STUDY00003882).


:::{.callout-tip title="Funding"}
All phases of this study were supported by a Childhood Arthritis and 
Rheumatology Research Alliance-Arthritis Foundation Transdisciplinary Research 
Grant (PI: Natoshia R. Cunningham).
:::

::: {.notes}
My case study is based on a forthcoming paper in *Arthritis Care & Research*.
Supporting references for things I'm discussing today can be found in the paper.
The full citation is at the end of the slides. These slides are available from 
the link shown at the bottom of the screen. I also want to note that the 
study was approved by MSU's institutional review board, and funded by the 
Childhood Arthritis and Rheumatology Research Alliance.

[In speaker notes, the symbol ▶ marks when to press the spacebar or click the 
right arrow to reveal the next animated element on a slide.]
:::

## Contributors {.smaller}

:::: {.columns}
::: {.column width="34%"}
* **Natoshia R. Cunningham**
* Thea Senger-Carpenter
* Jocelyn Zuckerman
* Michelle Adler
* Mallet R. Reid 
* Ashley Danguecan
* Luana Flores Pereira
* Sarah I. Mossad
* Samantha L. Ely
* Khalid Abulaban
:::

::: {.column width="33%"}
* Elizabeth A. Kessler
* Natalie Rosenwasser
* Tamar B. Rubinstein
* Ekemini A. Ogbu
* Emily A. Smitherman
* Alaina Miller
* Taylor Abounader
* Elizabeth Ross
* Livie Timmerman
* Dhriti Sharma
:::

::: {.column width="33%"}
* Jennifer N. Stinson
* Stacy Allen
* Kabita Nanda
* Tala El Tal
* Deborah M. Levy
* Linda T. Hiraki
* Hermine I. Brunner
* Mathew Reeves
* Steven J. Pierce
* **Andrea Knight**
:::

:::: 

**Bold = Principal investigators.**

::: {.notes}
Studies like the one I'm describing today can involve very large research teams
because recruiting participants from multiple clinics requires having team
members at each clinic. So, I want to just share the list of co-authors on our
paper to properly acknowledge everyone's contributions. The principal
investigators' names are in bold text here.
:::

# Introduction

::: {.notes}
With those preliminaries out of the way, I'll turn to introducing the study. 
:::

## Childhood-Onset Systemic Lupus Erythematosus (cSLE)

* A chronic, multisystemic, autoimmune disease.
  
::: {.incremental}
* Higher risk for depressive, fatigue, and pain symptoms, which may lower 
  health-related quality of life (HR-QOL).
* More often affects female and Asian, Native American, and Black youth.
* Black youth may get less help managing comorbid symptoms than White peers.

:::

::: {.notes}
Childhood-onset lupus is a chronic, multisystemic, autoimmune disease diagnosed 
prior to age 18 years. 

▶ Patients are at higher risk for depressive, fatigue, and pain symptoms, which
may lower health-related quality of life.

▶ We know that lupus has a disproportionate impact on some parts of the
population: it's more likely to affect female and Asian, Native American, and
Black youth. It's the fifth leading cause of death for Black and Hispanic
females age 15-24.

▶ Previous studies also suggest that Black youth may be less likely to receive
effective help managing comorbid symptoms such as anxiety and depression than
White peers. This all means that we need better interventions to help lupus
patients.
:::

## Cognitive Behavioral Therapy (CBT)

* Effective for managing depression, fatigue, and pain in youth with other 
  chronic illnesses and in *adults* with lupus. 

::: {.incremental}
* Prior studies on CBT for *youth* with lupus: 
    * One failed to find clinically significant effects on pain, affect, & 
      quality of life. 
    * Treatment and Education Approach for Childhood-Onset Lupus (TEACH) program 
      had promising results but needed refinements. 

:::

::: {.notes}
Psychological interventions offer one way to improve the mental health of lupus
patients. Research has shown that cognitive behavioral therapy (CBT) is
effective for managing depression, fatigue, and pain in youth with *other* 
*chronic illnesses* and in *adults* with lupus.

▶ However, there were only two prior studies on using CBT for youth with lupus. 

▶ One of them failed to find any clinically significant effects on pain, affect,
and quality of life.

▶ The other was our team's single-arm pilot test of the Treatment and Education
Approach for Childhood-Onset Lupus, the TEACH program. My colleagues' first
TEACH study yielded promising results but showed that it still needed some
refinements.
:::

## TEACH: Initial Pilot Study

* A tailored, CBT program delivered via 6 weekly sessions.
* Targeted depressive, fatigue, & pain symptoms.
* High retention rate (82.4%).
* Reduced depressive and fatigue symptoms.
* Marginal reductions in pain and anxiety. 

::: {.notes}
The first iteration of the TEACH intervention consisted of a tailored, CBT
program delivered during a series of 6 weekly in-person visits. The primary
outcomes it sought to improve were patients' depressive, fatigue, and pain
symptoms. The first study had a patient retention rate of just over 82% and
potential improvements in depressive and fatigue symptoms, with weaker evidence
for reductions in pain and anxiety.

Intervention development is often iterative. Early results like these are 
promising but not definitive and usually accompanied by feedback about ways to 
improve the program. That usually means we conduct multiple studies, starting 
with small scale studies in early iterations and using that feedback to fix 
problems in subsequent iterations, where testing efficacy and using larger 
samples become more important. So think of this case study as an example of that 
process. 
:::

## TEACH: Initial Lessons Learned

* Patients wanted a remote delivery option. 
* Need to engage patients in refining the intervention.
* Need to better address pain & anxiety. 
* Need to get stronger evidence of efficacy.

::: {.notes}
Some of the initial lessons learned were that patients reported wanting the
option to get the TEACH intervention delivered remotely instead of only through
in-person visits. That first study also revealed a need to refine the
intervention and it was clear that engaging lupus patients in planning those
refinements could be crucial to improving feasibility, acceptability, and
effectiveness of the intervention. One priority was better addressing pain and
anxiety. Finally, my colleagues also recognized that they needed a stronger
research design because the initial study lacked a control group. 
:::

## Current Study

* Refined TEACH intervention for adolescents/young adults with cSLE.
    * Update program content per previous feedback.
    * Switch to remote delivery via video calls. 
* Primary outcomes: feasibility & acceptability. 
* Test efficacy of TEACH vs. TAU:
    * Secondary: depressive, fatigue, & pain symptoms. 
    * Exploratory: disease activity, medication adherence, & anxiety symptoms.

::: {.notes}
That led to developing the study I'm describing today, which aimed to refine
the TEACH intervention for adolescents and young adults with lupus. The PIs
updated the program content and delivery method to address the patient feedback
from the prior study. That included switching to delivering the intervention
remotely via video calls with participants instead of in-person.

One major objective was to examine feasibility and acceptability as primary
outcomes of the updated program. The other major objective was to move to a
randomized controlled trial design that included an intervention arm and a
control arm. That design change aimed to improve internal validity when we
tested the intervention's efficacy by comparing secondary and exploratory
outcomes among patients who received TEACH with outcomes among those who did
not. Both groups got medical treatment-as-usual for lupus throughout the study,
so patients in the intervention arm received TEACH paired with usual treatment
and the control arm received only treatment as usual.

Measures of depressive, fatigue, and pain symptoms were the secondary outcomes, 
while measures of disease activity, medication adherence, and anxiety symptoms 
were exploratory outcomes. 
:::

# Methods

::: {.notes}
So, now I'll tell you about the methods we used in this study. 
:::

## Research Design

```{r}
#| results: asis

StudyDesign <- data.frame(Arm = c("TEACH", "TAU"),
                          Pretest = rep("$O_1$", times = 2),
                          Allocation = rep("$R$", times = 2),
                          Intervention = c("$X$", ""),
                          Posttest = rep("$O_2$", times = 2)) 

FN <- paste("$O_i$, observed data at time $i$;", 
            "$R$, randomized;",
            "$X$, delivery of TEACH intervention (6 weeks).")
kable(StudyDesign, format = "html", escape = FALSE, align = "lcccc",
      col.names = c("Arm", "Pretest Week 0", "Allocation", "Intervention", 
                    "Posttest Week 8")) %>% 
  kable_styling(full_width = FALSE) %>% 
  add_header_above(c(" ", "Longitudinal Event" = 4), escape = FALSE) %>% 
  footnote(general = FN, general_title = "Note: ", footnote_as_chunk = TRUE,
           threeparttable = FALSE, escape = FALSE) %>% 
  cat()
```

* Preregistered at ClinicalTrials.gov (NCT04335643)

::: {.notes}
This study used a longitudinal, parallel arms randomized controlled trial
design, with pretest data collected before random assignment. This table is a
simple way to visualize the study design. Each row represents one of the two
arms we want to compare. The columns correspond to different longitudinal events
that unfold over time as we move from left to right in the table. Collection of
observed data is denoted with an O in both the pretest and posttest columns,
while random assignment to arms is shown by the R in the allocation column.
Delivery of the intervention is shown by placing an X in the intervention
column.

We expect no differences on pretest outcomes because we collected those data
before assigning patients to the two arms. At that point they are all part of
the same population. After randomization, their experiences diverge
systematically. Because one arm goes on to get the TEACH intervention for 6
weeks and the other does not, collecting posttest data 8 weeks after pretest
enables us to test intervention efficacy. Using the pretest data as a covariate
increases statistical power because repeated measures of an outcome from the
same patient are often correlated over time. It also means our test of the arm
effect quantifies the expected difference in posttest outcomes among patients
who start with the same pretest score but end up in opposite arms.
:::

## Sites: Pediatric Rheumatology Clinics

:::: {.columns}
::: {.column width="50%"}
**United States**

* Helen DeVos Children's Hospital
* Cincinnati Children's Hospital Medical Center
* Seattle Children's Hospital
* Children's Hospital at Montefiore
* University of Alabama

:::

::: {.column width="50%"}
**Canada**

* The Hospital for Sick Children, Toronto

:::
::::

::: {.notes}
This was an international, multi-site sudy. Our sites were all pediatric
rheumatology clinics. We had five sites in the US and one in Canada.
:::

## Participants {.smaller}

:::: {.columns}
::: {.column width="50%"}
**Inclusion Criteria**

* Age 12-22 years, 
* cSLE diagnosis by age 18 yrs,
* Elevated symptoms:
    * Depressive: $60 \le T < 90$
    * Fatigue: $T \ge 60$
    * Pain: $\ge 3$
* English language proficiency. 

:::

::: {.column width="50%"}
**Exclusion Criteria**

* Other chronic medical conditions, 
* Developmental delays or severe cognitive impairments,
* Untreated major psychiatric illness, or
* In treatment for depressive, fatigue, or pain symptoms. 

:::
::::

::: {.notes}
Participants had to be 12-22 years old; with a lupus diagnosis from a pediatric 
rheumatologist by age 18 yrs; have elevated depressive, fatigue, or pain 
symptoms; and be proficient in English. We excluded any who had other chronic 
medical conditions (such as juvenile idiopathic arthritis); any documented 
developmental delays or severe cognitive impairments; untreated major 
psychiatric illness (such as bipolar disorder, active suicidal ideation); or
who was already getting treated for depressive, fatigue, or pain symptoms. 

Participants were compensated with gift cards: $50 each at baseline and 
post-assessment. 
:::

## TEACH Intervention

* Advisory input from cSLE patients & caregivers.
* A tailored, CBT program, 6 weekly video calls.
* Delivered by 1 of 3 interventionists.
* Supporting online modules, caregiver involved for minors
* Still targets depressive, fatigue, & pain symptoms.
* Psychoeducation, behavioral, cognitive, & mindfulness strategies. 

::: {.notes}
The TEACH intervention was adapted with input from an advisory group of lupus 
patients and caregivers. It remained a 6-session CBT program, but was delivered 
via video calls with trained interventionists (two licensed clinical social 
workers or a pediatric psychologist). Patients used supporting online modules, 
and caregivers were involved for minors. TEACH continued to target depressive, 
fatigue, and pain symptoms as key outcomes. The intervention content 
included psychoeducation, behavioral strategies such as relaxation training and 
activity pacing, cognitive strategies such as challenging automatic thoughts and 
problem solving, and mindfulness strategies. 
:::

## Fidelity Assessment

* Fidelity is the degree to which an intervention is delivered as intended by 
  its developers. 
* Low fidelity may weaken intervention effects. 
* Our team recorded all video sessions with patients.
* Evaluators scored fidelity of each session (0-100%) via standardized 
  checklist. 

:::: {.fragment}
::: {.callout-tip}
**Average fidelity was 98%!**
:::

::::

::: {.notes}
An intervention's fidelity is the degree to which it is delivered as intended by
its developers. For example, that may require following a treatment protocol
described in an intervention manual. High fidelity helps you ensure that study
results show the actual efficacy instead of the result of inconsistent usage or
implementation. Low fidelity may reduce or weaken intervention effects of
otherwise useful interventions.

Our team recorded video of all intervention sessions, then had three 
independent, trained evaluators (a physician, licensed social worker, or
doctoral student of psychology) use standardized checklists to assess fidelity 
to the TEACH intervention manual, yielding a score that ranged from 0 to 100%. 

▶ Fidelity was extraordinarily high: the average across all sessions was 98%!
:::

## Stratified Random Block Randomization

```{r}
#| results: asis

Strata <- data.frame(Stratum = 1:4,
                     Country = rep(c("United States", "Canada"), each = 2),
                     Age_Group = rep(c("Adolescent (12-17 yrs)", 
                                       "Young Adult (18-22 yrs)"), times = 2)) 

FN <- paste("Within stratum, 1:1 ratio for arms (random blocks, n = 4 or 8).",
            "Investigators were blind to randomization.",
            "TAU arm offered TEACH after posttest.")
kable(Strata, format = "html", escape = FALSE, align = "ccc",
      col.names = c("Stratum", "Country", "Age Group")) %>% 
  kable_styling(full_width = FALSE) %>% 
  footnote(general = FN, general_title = "Note: ", footnote_as_chunk = TRUE,
           threeparttable = FALSE, escape = FALSE) %>% 
  cat()
```

::: {.notes}
To balance sample size across arms, the study used a stratified random block
randomization process. The two stratification variables were country (US vs
Canada) and age group (adolescent vs young adult). Within each stratum, blocks
of either 4 or 8 participants were randomly allocated to the TEACH vs TAU arms
in 1:1 ratio. The investigators (except the interventionists & statistician)
were blind to randomization results during data collection. After the posttest,
the TAU participants were offered the TEACH intervention.

Using stratified randomization means we have to add the stratification variables 
to our model as covariates to control for confounding, reduce error variance, 
and improve precision of the intervention effect estimate. 
:::

## Primary Outcome Measures 

* Recruitment rate = (# consented)/(# approached)
    * Benchmark: > 65 %
* Retention rate = (# completed posttest)/(# randomized)
    * Benchmark: > 80%
* Qualitative interviews about TEACH acceptability

::: {.notes}
The primary quantitative outcomes for this study were the recruitment rate, 
defined as the percent of approached patients who consented to joining the 
study, and the retention rate, defined as the percent of randomized patients who 
completed the posttest data collection. We set a target benchmark for each rate 
before collecting data: We wanted to get a recruitment rate above 65% and a 
retention rate over 80%. 

Qualitative feedback about the acceptability of the intervention was collected
by interviewing participants and used to assess acceptability as another primary
outcome. That makes this a mixed methods study because we're drawing on both
quantitative and qualitative approaches.
:::

## Secondary Outcome Measures

* Depressive symptoms over past 2 weeks. 
    * CDI-II (28 items) or BDI-II (21 items), $T$-scores.
* Fatigue over past week.
    * PROMIS PSF/ASF (10/8 items), $T$-scores.
* Pain intensity over past 2 weeks.
    * Pain Visual Analog Scale (VAS), scores 0-10.
  
::: {.notes}
We had three secondary outcomes. We measured depressive symptoms with the 
Children's Depression Inventory 2nd Edition (CDI-II) for adolescents and Beck 
Depression Inventory-II (BDI-II) for young adults, converting both to T-scores 
with Mean = 50 and SD= 10 to make them comparable across age groups. A 50%
reduction is considered meaningful. 

Fatigue was measured by the PROMIS Pediatric or Adult Short Form, again
converting to T-scores for comparability. A fatigue reduction of 10 points is
meaningful.

Finally, pain was measured with the pain visual analog scale, which uses a 0 to 
10 scale and where a 30% reduction is meaningful.
:::

## Exploratory Outcome Measures {.smaller}

* Anxiety for past 3 months
    * SCARED/SCAARED (41/44 items), $z$-scores.
* Disease activity
    * SLEDAI-2K, scores from 0-105 ($\ge 6$ shows active disease). 
* Medication adherence over past month
    * Adapted from Medication Adherence Self-Report Inventory, 0-100%
    * Scores $\ge 80$% indicate adequate adherence.
* Pediatric health-related quality of life (HR-QOL)
    * PedsQL Quality of Life Inventory (child & caregiver)
    * PedsQL Rheumatology Module (child & caregiver)

::: {.notes}
We also had several exploratory outcomes. We measured anxiety with the Screen
for Child Anxiety Related Disorders (SCARED) for adolescents and the Screen for
Adult Anxiety Related Disorders (SCAARED) for young adults, converting both to
z-scores for comparability. Reductions of over 50% are meaningful.

We used the Systemic Lupus Erythematosus Disease Activity Index 2000 (SLEDAI-2K)
to measure disease activity. SLEDAI scores range from 0-105, with scores $\ge 6$
indicating active disease.

Medication adherence over past month was measured with an instrument adapted
from the Medication Adherence Self-Report Inventory. Scores are rated 0-100%,
with $\ge 80$% indicating adequate adherence.

We used some measures of health-related quality of life as well, but I'll omit
results for those today to keep the presentation focused.
:::

## Other Measures

* Demographics: Age, sex, race, ethnicity, disease duration (months), education.
* Disease manifestations: Systemic Lupus International Collaborating Clinics 
  (SLICC) classification.
* Adverse childhood experiences (ACEs, scores 0-9).

::: {.notes}
Naturally, we collected other measures too. For demographics, we had age, sex, 
race, ethnicity, disease duration (in months), and caregiver education. We used 
Systemic Lupus International Collaborating Clinics (SLICC) classification to 
measure disease manifestation, and we had a measure of the number of adverse 
childhood experiences for each patient. 

Some of these measures were collected just to describe the sample, others were 
used in our models. 
:::

## Qualitative Analysis

* Grounded theory & thematic analytic framework.
* Interviews recorded & transcribed.
* Reviewed by 3 trained coders.
* Quotes coded into themes until saturation & consensus achieved.

::: {.notes}
My colleagues used a grounded theory and thematic analytic framework to analyze 
the qualitative data. Participant interviews were recorded, transcribed, and 
reviewed by 3 trained coders. They coded quotes into a set of a priori and 
emerging themes until they achieved both saturation and consensus. 
:::

## Quantitative Analysis

* Estimated recruitment and retention rates.
* Intent-to-treat (ITT) estimate of intervention effect for secondary & 
  exploratory outcomes.
    * Use all randomized participants.
    * Stay in assigned arm regardless of non-compliance, missing data, etc.
    * Gold standard for testing efficacy in RCTs.

::: {.notes}
For the primary outcomes, the quantitative analysis was simple: we estimated the
recruitment and retention rates and compared them to our benchmarks. 

For secondary and exploratory outcomes, we wanted to get intent-to-treat
estimates of the intervention effect. That means that every randomized patient
has to stay in the analysis and in the arm to which they were assigned,
regardless of non-compliance, withdrawal, missing data, or protocol deviations.
So a TEACH patient who skipped half of the planned intervention sessions was
still treated as a full member of the TEACH arm even though they didn't get the
full dosage. 

ITT estimates are the gold standard for analyzing randomized trials to get a
pragmatic, real-world estimate of efficacy in part because they represent what
happens when we prescribe a treatment in settings where compliance is imperfect
instead of what happens when everyone fully complies with the treatment.
:::

## Analysis of Covariance (ANCOVA)

```{dot}
//| fig-width: 10
//| fig-height: 6
//| 
digraph ANCOVA {
rankdir="LR";

node [shape = "box", style= "filled", fontname="Lato", fillcolor = "#C3FFEC"]
Post [label = "Posttest Measure\n(Dependent Variable)"]
Arm [label = "Arm\n(Focal Predictor)"]
Pre [label = "Pretest Measure\n(Key Covariate)"]
ArmxPre [label = "Arm x Pretest\n(Interaction)"]
Covariates [label = "Country, Age Group, Race, Ethnicity,\nACEs (Other Covariates)"]

Arm -> Post 
Pre -> Post
ArmxPre -> Post
Covariates -> Post
}
```    

::: {.notes}
To get our ITT estimates, we used the posttest outcomes as the dependent variable
in analysis of covariance (ANCOVA) models and tested for an arm effect,
controlling for pretest scores on the outcome measure. We also examined whether
pretest scores moderated the intervention effect by testing an arm by pretest
interaction. 

We included several other covariates, but sample size limited how many we could
use while still getting trustworthy estimates. We had to control for the
stratification covariates of country and age group because of the randomization
design. We also controlled for race, Hispanic ethnicity, and adverse childhood
experiences. However, the distribution of sex was so extreme that we could not
use it as a covariate. 

We use estimated marginal means, contrasts between them, and some graphs to
interpret the intervention effects.
:::

## Imputing Missing Data via MICE {.smaller}

* Allowed retaining all cases as required for ITT analysis 
* 55% of cases had at least one missing value.
* Imputation model used:
    * All focal predictors & covariates used in the (ANCOVA) model.
    * Binary indicator of post-assessment attendance.
    * Longitudinal relationships between pretest & posttest for each outcome.
    * Cross-sectional relationships among outcomes measured at the same time. 
* Parameter estimates pooled across analyses of 55 imputed data sets. 

::: {.notes}
Intent-to-treat analyses require retaining all randomized participants in the 
analysis. By default most statistical software will drop cases with missing data 
on any of the variables involved in a model. That can cause serious biases, so 
we handled missing data with multiple imputation via chained equations (often 
abbreviated as MICE). This method replaces each missing value by a set of draws 
from the conditional distribution of that variable informed by the imputation 
predictors, yielding not just one completed dataset but a whole collection of 
imputed data sets that we have to analyze to get better estimates. 

About 55% of the cases in our data set had at least one missing value, largely
because the quality of life measures were missing by design for young adult
patients. Those measures were only developed for use with adolescents. Our
imputation model included all the focal predictors & covariates used in the
(ANCOVA) model, plus a binary indicator of post-assessment attendance,
longitudinal relationships between pretest & posttest for each outcome, and
cross-sectional relationships among outcomes measured at the same time.

We generated 55 imputed datsets, analyzed each one with the same model, then 
pooled the parameter estimates across data sets.
:::

# Results

::: {.notes}
So, now let's see some of the results. 
:::
## Sample Descriptives

* Country: 61% US, 39% CA
* Age: M = 16.5, SD = 2.1, 64% adolescent, 36% young adult
* Sex: 97% female, 3% male
* Race: 42% White, 25% Asian, 22% Black, 9% Other, 2% Missing
* Ethnicity: 77% Not Hispanic, 22% Hispanic, 2% Missing

::: {.notes}
Here is an overall demographic summary of the participants. About 61% were
from the US, with the rest from Canada. The mean age was 16.5 years;
64% of the sample were adolescents and 36% were young adults. About 97% were
female and about 42% were white, with 25% Asian, and 22% Black. About 22% were 
Hispanic. 
:::

## CONSORT Chart

```{dot}
//| fig-width: 10
//| fig-height: 2.5
digraph CONSORT {
rankdir="LR";

node [shape = "box", style= "filled", fontname="Lato", fillcolor = "#C3FFEC"]
Identified [label = "Identified\nn = 293"]
Approached [label = "Approached\nn = 200"]
Consented [label = "Consented\nn = 97"]
Screened [label = "Screened\nn = 85"]
Pretest [label = "Pretest &\nRandomized\nn = 64"]
TEACH [label = "TEACH\nn = 32"]
TEACHPosttest [label = "Posttest\nn = 29"]
TAU [label = "TAU\nn = 32"]
TAUPosttest [label = "Posttest\nn = 30"]

Identified -> Approached -> Consented -> Screened -> Pretest 
Pretest -> TEACH -> TEACHPosttest
Pretest -> TAU -> TAUPosttest
}
```

:::: {.fragment}
:::{.callout-tip title = "Primary Outcomes"}

* Recruitment rate = 97/200 = 48.5% (benchmark: 65%)
* Retention rate = 59/64 = 92.2% (benchmark: 80%)
    * Unrelated: arm, country, age group, race, ethnicity, most pretest 
      outcomes. 
    * Related: pretest child rheumatology QOL, disease duration.

:::
::::

::: {.notes}
This CONSORT chart summarizes participant flow through the study. It started by 
identifying potential participants, then approaching them to ask for consent. 
Then, those who consented were screened and the eligible participants then took
the pretest and were randomly assigned to one of the two arms (TEACH vs TAU). 
Then, 8 weeks later, the posttest assessment occurred. The chart in our paper 
shows breakdowns of why people dropped out between steps. Here, I want 
to emphasize that we can use the data shown here to estimate two of the primary 
outcomes, the recruitment rate and the retention rate. 

▶ 97 of 200 patients approached consented to participate, yielding a 48.5%
recruitment rate. That was lower than the 65% benchmark we hoped to achieve.
However, 59 of 64 randomized participants returned for the posttest assessment,
giving us a 92.2% retention rate that exceeded the 80% benchmark we had set. In
our attrition analyses, retention at posttest was unrelated to arm, country, age 
group, race, ethnicity, and most pretest outcomes, but pretest child 
rheumatology quality of life and disease duration were related to retention.
:::

## Pretest Symptoms (Raw/Unimputed)

```{r}
#| results: asis

PreSymptoms <- data.frame(Symptom = c("Depressive Symptoms",
                                      "Fatigue",
                                      "Pain Intensity",
                                      "Anxiety"),
                          TEACH = c("63.6 (12.0)",
                                    "61.6 (8.5)",
                                    "3.5 (2.0)",
                                    "0.1 (1.1)"),
                          TAU = c("64.4 (10.4)",
                                  "60.1 (6.7)",
                                  "4.0 (1.9)",
                                  "-0.1 (0.8)"))

kable(PreSymptoms, format = "html", escape = FALSE, align = "lrr",
      col.names = c("Variable", "TEACH M (SD)", "TAU M (SD)")) %>% 
  kable_styling(full_width = FALSE)
```
::: {.notes}
Before we look at the intervention effect, we should the pretest data about the
secondary and exploratory outcomes. This table shows pretest means and standard
deviations of the raw, unimputed data broken down by arm. One thing to notice
here is that the means are very similar between arms: that's what we expect
because random assignment occurred after these data were collected.

We can also see that the mean T-scores for depressive symptoms and fatigue were 
just above their respective eligibility criteria, both of which required scores 
of at least 60. Similarly, the pain eligibility criterion was a score of at 
least 3 and the means here are 3.5 and 4. Remember that a participant could 
qualify due to an elevated score on any of those three symptoms: we didn't 
require them to be high on all of them. 
:::

## Pretest cSLE-Related Factors (Raw/Unimputed)

```{r}
#| results: asis

PreLupus <- data.frame(Symptom = c("Months since diagnosis",
                                   "No. of Medications",
                                   "Medication Adherence",
                                   "Disease Manifestations",
                                   "Disease Activity"),
                          TEACH = c("*31 (1-192)*",
                                    "*2 (0-12)*",
                                    "83.7 (15.7)",
                                    "0.50 (0.75)", 
                                    "*5 (0-18)*"),
                          TAU = c("*26 (0-178)*",
                                  "*2 (0-10)*",
                                  "84.1 (19.2)",
                                  "0.53 (1.0)",
                                  "*2 (0-12)*"))
FN <- "Italics: *median (range)*. Normal font: mean (SD)"

kable(PreLupus, format = "html", escape = FALSE, align = "lrr",
      col.names = c("Variable", "TEACH", "TAU")) %>% 
  kable_styling(full_width = FALSE, font_size = 38) %>% 
  footnote(general = FN, general_title = "Note: ", footnote_as_chunk = TRUE,
           threeparttable = FALSE, escape = FALSE) %>% 
  cat()
```

::: {.notes}
This table shows pretest data by group for some of the lupus-related factors 
and exploratory outcomes. Again, we see similar values between arms, consistent 
with what we expect due to randomization. The median time since lupus diagnosis 
was 26 to 31 months. The median number of medications was 2 in both arms. The 
most important finding on this slide is that the means for medication adherence 
were 83-84% at pretest in both groups, suggesting adherence was already adequate 
and that there's not a lot of room for improvement. Keep that in mind. 
:::

## Primary Outcomes: Acceptability Themes

* Skills were logically connected & went "hand in hand".
* Relaxation skills, pleasant activity planning, & activity pacing were helpful.
* There was immediate effect of using the skills taught.
* Relationship w/ interventionist motivated practicing skills.
* Symptom severity & busy schedules were barriers.
* Virtual format was appreciated & facilitated attendance. 

::: {.notes}
The themes emerging from our qualitative analyses suggest that the patients
thought the skills emphasized by TEACH were logically connected and went hand in 
hand. They reported that the relaxation skills, pleasant activity planning, and 
activity pacing strategies were helpful, and that there was an immediate benefit 
of using the skills they were taught. Patients also said that having a positive 
relationship with the interventionist motivated them to actually practice using 
the skills (even though they struggled to remember the names of the skills). 
Greater symptom severity and busy schedules were identified as barriers to using 
TEACH skills, but patients enjoyed the virtual format and said it facilitated 
attendance. 
:::

## Secondary Posttest Outcomes: Arm Main Effect

```{r}
#| fig-height: 1.5
#| fig-width: 5
#| layout-ncol: 2

plot(emm_mood, ylab = NULL, xlab = "Depressive symptoms, Posttest Means") +
  scale_x_continuous(limits = c(40, 70), breaks = seq(40, 70, 10))

plot(emmc_CI_mood, ylab = NULL, xlab = "Depressive symptoms, Posttest Difference") + 
  scale_x_continuous(limits = c(-15, 15), 
                     breaks = seq(from = -15, to = 15, by = 5)) +
  geom_vline(xintercept = 0, col = "red") 

plot(emm_fatigue, ylab = NULL, xlab = "Fatigue, Posttest Means") +
  scale_x_continuous(limits = c(40, 70), breaks = seq(40, 70, 10))

plot(emmc_CI_fatigue, ylab = NULL, xlab = "Fatigue, Posttest Difference") + 
  scale_x_continuous(limits = c(-15, 15), 
                     breaks = seq(from = -15, to = 15, by = 5)) +
  geom_vline(xintercept = 0, col = "red") 

plot(emm_pain, ylab = NULL, xlab = "Pain, Posttest Means") +
  scale_x_continuous(limits = c(1, 5), breaks = seq(1, 5, 0.5))

plot(emmc_CI_pain, ylab = NULL, xlab = "Pain, Posttest Difference") + 
  scale_x_continuous(limits = c(-2, 2), 
                     breaks = seq(from = -2, to = 2, by = 1)) +
  geom_vline(xintercept = 0, col = "red") 
```

::: {.notes}
Now let's look at the results for intervention efficacy. These are based on the
imputed data and are adjusted by averaging over other covariates. Here are the
results for testing the main effect of arm, which tells us whether on average
the TEACH arm did better than the TAU arm at posttest. On all these secondary
outcomes, lower scores are desirable. On the left, you can see the posttest
means and 95% confidence intervals for each arm on each secondary outcome. The
posttest means for *both arms* are notably lower than the pretest means I showed 
earlier. But what matters for efficacy is how these posttest means compare to 
each other. 

So, on the right, I've graphed the raw effect size measured by the
difference between those means. When the red reference line intersects the
confidence interval for the difference, that means there may be no difference.
Here we see that on average, depressive symptoms improved by almost 8 points
(14%) and fatigue by almost 4 points (7%). Pain, however may not have improved
despite an average reduction of 0.89 points (25%).
:::

## Fatigue: Arm x Pretest Interaction

```{r}
#| fig-width: 10
#| fig-height: 5.5

emmip(fit_fatigue_m3, formula = arm ~ c_fatigue_T1, cov.reduce = range, 
      style = "numeric", CIs = TRUE, CIarg = list(lwd = 1, alpha = 0.25), 
      linearg = list(lwd = 1.2), 
      xlab = "Centered Pretest Fatigue",
      ylab = "Posttest Fatigue") + 
  geom_ribbon(aes(ymin=LCL, ymax=UCL, fill=arm), alpha=.25) + 
  theme(text = element_text(family = "sans", size = 14),
        axis.text = element_text(family = "sans", size=14),
        legend.position = "inside", 
        legend.position.inside = c(0.85, 0.20),
        legend.title = element_text(family = "sans", size=12),
        legend.text = element_text(family = "sans", size=12)) +
  coord_cartesian(xlim = c(-31, 21), ylim = c(20, 71))
```

::: {.notes}
Testing the arm by pretest interaction for fatigue showed that the size of the
intervention effect depended on pretest levels of fatigue. Here, the blue line
represents the TAU arm and the orange one is the TEACH arm. The vertical
distance between the lines is the intervention effect at a given pretest level
of fatigue. The lines are not parallel because there is an interaction. TEACH
has the strongest beneficial effect on posttest fatigue when pretest fatigue is
very mild (> 10 points below average). The benefit disappears by about 7.5
points above average pretest fatigue. At unusually high levels of pretest
fatigue (> 10-12 points above average), TEACH appears to worsen posttest
fatigue.
:::

## Exploratory Posttest Outcomes: Arm Main Effect

```{r}
#| fig-height: 1.5
#| fig-width: 5
#| layout-ncol: 2

plot(emm_anxiety, ylab = NULL, xlab = "Anxiety, Posttest Means") +
  scale_x_continuous(limits = c(-0.75, 0.75), breaks = seq(-0.75, 0.75, 0.25))

plot(emmc_CI_anxiety, ylab = NULL, xlab = "Anxiety, Posttest Difference") + 
  scale_x_continuous(limits = c(-0.75, 0.75), 
                     breaks = seq(from = -0.75, to = 0.75, by = 0.25)) +
  geom_vline(xintercept = 0, col = "red") 

plot(emm_sledai, ylab = NULL, xlab = "Disease Activity, Posttest Means") +
  scale_x_continuous(limits = c(1, 10), breaks = seq(1, 10, 1))

plot(emmc_CI_sledai, ylab = NULL, xlab = "Disease Activity, Posttest Difference") + 
  scale_x_continuous(limits = c(-4, 4), 
                     breaks = seq(from = -4, to = 4, by = 1)) +
  geom_vline(xintercept = 0, col = "red") 

plot(emm_masri, ylab = NULL, xlab = "Medication Adherence, Posttest Means") +
  scale_x_continuous(limits = c(80, 110), breaks = seq(80, 110, 10))

plot(emmc_CI_masri, ylab = NULL, xlab = "Medication Adherence, Posttest Difference") + 
  scale_x_continuous(limits = c(-10, 10), 
                     breaks = seq(from = -10, to = 10, by = 5)) +
  geom_vline(xintercept = 0, col = "red") 
```

::: {.notes}
Here are results for some of the exploratory outcomes. We can see that the TEACH
arm ended up with less anxiety than the TAU arm, by about 0.35 points on the 
z-score scale. There wasn't any effect on disease activity.

Now, remember that pretest medication adherence was already adequate and very
similar across arms, averaging about 83-84%. Here we can see that by the 8-week
posttest it had increased to about 92-93% for both arms, which is great. We want
to see high adherence. But, it improved by a similar amount in each arm, so it's
not surprising that we don't see an overall intervention effect on adherence at
posttest. However, that is only looking at part of the story. We actually found
something more interesting than that.
:::

## Medication Adherence: Arm x Pretest Interaction

```{r}
#| fig-width: 10
#| fig-height: 5.5

emmip(fit_masri_m3, formula = arm ~ c_masri_score_T1, cov.reduce = range, 
      style = "numeric", CIs = TRUE, CIarg = list(lwd = 1, alpha = 0.25), 
      linearg = list(lwd = 1.2), 
      xlab = "Centered Pretest Medication Adherence", xlim = c(-81, 41),
      ylab = "Posttest Medication Adherence") + 
  geom_ribbon(aes(ymin=LCL, ymax=UCL, fill=arm), alpha=.25) + 
  theme(text = element_text(family = "sans", size = 14),
        axis.text = element_text(family = "sans", size=14),
        legend.position = "inside", 
        legend.position.inside = c(0.85, 0.25),
        legend.title = element_text(family = "sans", size=12),
        legend.text = element_text(family = "sans", size=12)) +
  coord_cartesian(xlim = c(-60, 20), ylim = c(40, 120))
```

::: {.notes}
While there was no main effect on medication adherence, we did find an arm by
pretest adherence interaction. That means the size of the intervention effect
was not constant. Again the blue line is the TAU arm and the orange one is the
TEACH arm. Here, high adherence is good. We can see that TEACH has the strongest
beneficial effect on posttest medication adherence when pretest adherence is
very low (> 20 points below average). The benefit disappears at average pretest
adherence. At higher levels of pretest adherence (> 10 points above average),
TEACH may worsen posttest adherence.
:::

## Conclusions

* TEACH may improve depressive, fatigue, and anxiety symptoms
    * Need to look at longer time period to see if benefits persist. 
    * Improvements fell short of clinically significant effect sizes. 
    * Effect on fatigue depends on pretest. 
    * Pain started pretty low, so had little room to improve. 
* TEACH improved medication adherence among those with low pretest adherence. 
* Low recruitment rate suggests systemic barriers to care for this population. 
  We need to find ways to overcome those. 
* Patients appreciated the program, remote delivery, and relationship with 
  interventionist.
* 

## Limitations

* Small sample limits power & risks reduced representativeness. 
* Exclusion criteria limit generalizability. 
* TAU practices may vary across sites. 
* Self-reported medication adherence could be biased. 

::: {.notes}
Now, I want to acknowledge some of the limitations of our study. First, small
sample size limited the power and precision of estimates and comes with a risk
of reduced sample representativeness. Replication with larger samples will be
important for building the evidence base about how well TEACH actually works.

Second, our eligibility criteria excluded youth with more severe mental health
problems and/or disease activity and also those already in psychological
treatment. That may have affected the size of TEACH treatment effects, given the
greater potential room for improvement (as per our findings for medication
adherence). It may also affect generalizability of the results. 

Third, features of treatment as usual may vary between clinical sites. Future
work may include more detailed assessments of both TAU and other adjunctive
therapies used by youth and families.

Finally, self-reported measures of medication adherence are vulnerable to 
response bias. Future studies should consider more comprehensive assessment 
approaches.
:::

## Next Steps

A larger follow-up study is underway using a hybrid implementation-effectiveness 
design. 

## References {.scrollable}

::: {#refs}
:::

::: {.notes}
Finally, here is the reference for our paper. 
:::

# Questions & Discussion Time
Thank you for attending!

::: {.notes}
*Time should be ~40-45 min, leaving 5 min for questions and discussion.*
:::
